---
title: "BIOSTAT 625 Final Project"
author: "Salina Moon"
output:
  pdf_document: default
fontsize: 12pt
geometry: margin=1in
---
# Exploratory Data Analysis  
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)

data.imp = read.csv("merge_df.csv")
```

### First look at the data
```{r}
freq.stateyr = data.frame(unclass(table(data.nomiss$State,
                                        data.nomiss$Year)))
dim(freq.stateyr)
```
There are 51 states, including DC, and 3 time points: 2014, 2016, and 2018.

### Missingness
```{r}
# extract states with different numbers of counties over time
freq.stateyr %>% 
  mutate(same = (X2014==X2016)*(X2014==X2018)) %>%
  filter(same == 0)
```
The number of counties reporting these health information in each state only differs by 1, and in only 3 states. 
```{r}
data.nomiss = na.omit(data.imp)

c(nrow(data.imp), nrow(data.nomiss))
```
We only lose 8 observations out of over 9,000 if we only look at complete cases. 

### Change over time
#### State level
```{r}
# state average and dispersion
avg.state = data.nomiss %>%
  group_by(State) %>%
  summarise(Counties = n(),
            Diab = mean(Diagnosed.Diabetes.Percentage),
            SVI = mean(Overall.SVI),
            Food = mean(Food.Insecurity),
            Ins = mean(No.Health.Insurance),
            sdDiab = sd(Diagnosed.Diabetes.Percentage),
            sdSVI = sd(Overall.SVI),
            sdFood = sd(Food.Insecurity),
            sdIns = sd(No.Health.Insurance)) %>%
  mutate(Counties = round(Counties / 3)) %>%
  mutate(cvDiab = sdDiab / Diab) %>%
  mutate(cvSVI = sdSVI / SVI) %>%
  mutate(cvFood = sdFood / Food) %>%
  mutate(cvIns = sdIns / Ins) 

write.csv(avg.state, "bios625_avgstate.csv", row.names = FALSE)
```

#### County level
```{r}
# mean over all three time points
avg.county = data.nomiss %>%
  group_by(County_FIPS, State) %>%
  summarise(Diab = mean(Diagnosed.Diabetes.Percentage),
            SVI = mean(Overall.SVI),
            Food = mean(Food.Insecurity),
            Ins = mean(No.Health.Insurance),
            sdDiab = sd(Diagnosed.Diabetes.Percentage),
            sdSVI = sd(Overall.SVI),
            sdFood = sd(Food.Insecurity),
            sdIns = sd(No.Health.Insurance))

head(avg.county)
```
It is important to note that Food Insecurity and No Health Insurance have SD of 0 for all counties, meaning these two variables remain constant over time. Only Diabetes and SVI change between time points. 

Let's look at how SVI changes over time in each state. 
```{r}
yr.state = data.nomiss %>%
  group_by(State, Year) %>%
  summarise(nCounty = n(),
            Diab = mean(Diagnosed.Diabetes.Percentage),
            SVI = mean(Overall.SVI),
            Food = mean(Food.Insecurity),
            Ins = mean(No.Health.Insurance))

write.csv(yr.state, "bios625_state_byyear.csv", row.names = FALSE)

yr.state = yr.state[order(yr.state$State),]
yr.state = yr.state[order(yr.state$Year),]
svi.state.2014 = rep(yr.state[which(yr.state$Year==2014),]$SVI,3)
yr.state$deltaSVI = yr.state$SVI - svi.state.2014

ggplot(yr.state,
       aes(x = Year,
           y = deltaSVI,
           group = State)) +
  geom_line(color = "#FCAE91") +
  geom_point(aes(size = nCounty),
             color = "#DE2D26") +
  scale_x_continuous(breaks = c(2014, 2016, 2018)) +
  scale_y_continuous(breaks = seq(-0.06, 0.06, 0.02)) +
  scale_size_continuous(range = c(0.5, 3)) +
  labs(y = "Change in SVI from Baseline",
       size = "Counties") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"))

ggsave("bios625_time_svi.png", height=3, width=4.5)
```

### Correlation of Diabetes and Covariates
#### Diabetes vs. SVI
```{r}
lm(Diab ~ SVI, data = avg.county)
cor.test(avg.county$Diab, avg.county$SVI)

plot.diab.svi = ggplot(data.avgyr,
       aes(x = SVI,
           y = Diab)) +
  geom_point(color = "#FCAE91",
             size = 1) +
  geom_abline(slope = 2.748,
              intercept = 7.139,
              color = "#DE2D26",
              linewidth = 1) +
  geom_text(x = 0.5,
            y = 17,
            label = "R = 0.52, p<1E-15") +
  labs(y = "Diabetes (% Diagnosed)") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"))
```

#### Diabetes vs. Food Security
```{r}
lm(Diab ~ Food, data = data.avgyr)
cor.test(data.avgyr$Diab, data.avgyr$Food)

plot.diab.food = ggplot(data.avgyr,
       aes(x = Food,
           y = Diab)) +
  geom_point(color = "#BAE4B3",
             size = 1) +
  geom_abline(slope = 0.1714,
              intercept = 6.2675,
              color = "#31A354",
              linewidth = 1) +
  geom_text(x = 16.5,
            y = 17,
            label = "R = 0.42, p<1E-15") +
  labs(y = "Diabetes (% diagnosed)",
       x = "Food Insecurity") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())
```

#### Diabetes vs. No Health Insurance
```{r}
lm(Diab ~ Ins, data = data.avgyr)
cor.test(data.avgyr$Diab, data.avgyr$Ins)

plot.diab.ins = ggplot(data.avgyr,
       aes(x = Ins,
           y = Diab)) +
  geom_point(color = "#BDD7E7",
             size = 1) +
  geom_abline(slope = 0.03834,
              intercept = 8.14343,
              color = "#3182BD",
              linewidth = 1) +
  geom_text(x = 24,
            y = 17,
            label = "R = 0.13, p<1E-12") +
  labs(y = "Diabetes (% diagnosed)",
       x = "No Health Insurance") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())

ggsave("bios625_dot_diab_3var.png",
       grid.arrange(plot.diab.svi, plot.diab.food, plot.diab.ins,
                    ncol = 3,
                    widths = c(7,6,6)),
       height=3.5, width=5.72)
```

### National average
```{r}
data.nationyr = data.nomiss %>%
  group_by(Year) %>%
  summarise(Diab = mean(Diagnosed.Diabetes.Percentage),
            SVI = mean(Overall.SVI),
            Food = mean(Food.Insecurity),
            Ins = mean(No.Health.Insurance),
            sd.Diab = sd(Diagnosed.Diabetes.Percentage),
            sd.SVI = sd(Overall.SVI),
            sd.Food = sd(Food.Insecurity),
            sd.Ins = sd(No.Health.Insurance))
```
