---
title: "BIOSTAT 625 Final Project"
author: "Salina Moon"
output:
  pdf_document: default
fontsize: 12pt
geometry: margin=1in
---
# Exploratory Data Analysis  
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(scales)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)

data.imp = read.csv("merged_biofactor.csv")
```

### Missingness
```{r}
miss.inact = data.imp[is.na(data.imp$inactive),]
table(miss.inact$State)
```
There are missing data from AK, NJ, PR, VA. Let's evaluate the magnitude of missingness. 
```{r}
# subset Puerto Rico
state.pr = data.imp %>% filter(State == "Puerto Rico")

# count missingness
apply(state.pr, 2, function(x) sum(is.na(x))) / nrow(state.pr)
```
All covariate data are missing for Puerto Rico, so this state will be excluded moving forward. 

```{r}
# subset Alaska
state.ak = data.imp %>% filter(State == "Alaska")

# count missingness
state.ak %>%
  group_by(Year) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)

state.ak %>%
  group_by(County_FIPS) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)
```
58 (11%) of observations from Alaska are missing. Missingness is consistent across the years, in the same 8 counties. 
```{r}
# subset New Jersey
state.nj = data.imp %>% filter(State == "New Jersey")

# count missingness
state.nj %>%
  group_by(Year) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)

state.nj %>%
  group_by(County_FIPS) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)

# number of counties
length(unique(state.nj$County_FIPS))
```
All 21 missing observations occurred in 2019, from all 21 counties in New Jersey. 
```{r}
# subset Virginia
state.va = data.imp %>% filter(State == "Virginia")

# count missingness
state.va %>%
  group_by(Year) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)

state.va %>%
  group_by(County_FIPS) %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  filter(diabetes != 0)
```
Missingness in Virginia is fully accounted for from only one county between 2013 to 2019.
```{r}
data.nomiss = na.omit(data.imp)
```

### Summary statistics
```{r}
# distribution
hist(data.nomiss$inactive)
hist(data.nomiss$diabetes)
hist(data.nomiss$obesity)
```
All three variables approximate normal distributions. Diabetes can be appropriately modeled with linear regression, and inactivity and obesity do not need to be transformed. 

### Mean change over time
```{r}
# country-wide mean change over time
avg.country = data.nomiss %>%
  group_by(Year) %>%
  summarise(avginact = mean(inactive),
            avgdiab = mean(diabetes),
            avgobes = mean(obesity),
            sdinact = sd(inactive),
            sddiab = sd(diabetes),
            sdobes = sd(obesity)) %>%
  mutate(CVinact = sdinact / avginact) %>%
  mutate(CVdiab = sddiab / avgdiab) %>%
  mutate(CVobes = sdobes / avgobes)

write.csv(avg.country, "bios625_country.csv")

# convert to long format
avg.country.long = rbind(
  avg.country %>% 
    mutate(scaleinact = avginact / 2.5) %>% 
    select(Year, USMean = scaleinact) %>% 
    mutate(Variable = "Inactivity"),
  avg.country %>% 
    select(Year, USMean = avgdiab) %>% 
    mutate(Variable = "Diabetes"),
  avg.country %>% 
    mutate(scaleobes = avgobes / 2.5) %>% 
    select(Year, USMean = scaleobes) %>% 
    mutate(Variable = "Obesity")
)

# 3 variables vs Year
ggplot(data = avg.country.long,
       aes(x = Year,
           y = USMean,
           color = Variable)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2004, 2019, 5)) +
  scale_y_continuous(name = "Diabetes (%)",
                     sec.axis = sec_axis(trans = ~.*2.5, 
                                         name = "Inactivity or Obesity (%)",
                                         breaks = seq(18, 29, 3))) +
  labs(color = "") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"))

ggsave("bios625_time_3var.png", height=4, width=6)
```

```{r}
# state-wide mean over time
avg.state = data.nomiss %>%
  group_by(State, Year) %>%
  summarise(avginact = mean(inactive),
            avgdiab = mean(diabetes),
            avgobes = mean(obesity))

# Diabetes vs Year
ggplot(data = avg.state,
       aes(x = Year,
           y = avgdiab,
           group = State)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2004, 2019, 5)) +
  scale_y_continuous(breaks = seq(5, 13, 2)) +
  labs(y = "Prevalence of Diabetes (%)") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.length = unit(0.1, "inch"))

ggsave("bios625_time_state_var.png", height=4, width=6)
```

### Individual change over time
```{r}

```

